How to publish in Biological Sciences: a guide for the uninitiated

### 2021-12-15 1.09-preprints.Rmd
Add files via upload
# with lots of changes following copy edits and proofs

So far in this book, I have explained about the process of [peer review](#peerreview) how important this is and later I explain why it's the [silver](#fixing) (and not gold) standard in science. Preprints came to us from the world of physics. An academic world that is moving so quickly that many inside it don't want to wait for peer review before making their work public. They date back to 1991, and were the brainchild of Paul Ginsparg with his preprint server, arXiv. Today, arXiv hosts nearly 2 million articles in 8 subject areas: Physics, Mathematics, Computer Science, Quantitative Biology, Quantitative Finance, Statistics, Electrical Engineering and Systems Science and Economics.

@@

In an analysis by Abdill and Blekhman [-@abdill2019tracking] back in 2018, there were 37648 preprints uploaded to bioRxiv. But in the world of preprints, this information is quite dated as back then there were only ~2000 uploads each month. Luckily, these same authors regularly scrape data regarding submissions to bioRxiv and I'm able to reproduce these live data below. It is interesting to look back at the concern back in 2015 that preprint servers like bioRxiv might not catch on in the biological sciences [@vale2015accelerating], and it is still true to say that the number of preprints in the life sciences is still dwarfed by the annual number of publications, whereas physics has seen the opposite trend.

@@

What you see is a doubling of bioRxiv submissions dating back to May 2020 (Figure \@ref(fig:bioRxiv-line-annotated)), a few months after the start of the global COVID pandemic. (Go to the [RXivist website](https://rxivist.org/stats) to see the incredible spike in [medRxiv](https://www.medrxiv.org/) data at the same point in time.) Many scientists had spent a few months working from home. Some had been productive, and many decided to move this productivity for the first time onto bioRxiv.

@@

The preprint revolution has not gone unnoticed by the tech giants. Back in 2017 the [Chan Zuckerberg Initiative](https://chanzuckerberg.com/) made an undisclosed donation to bioRxiv. The advisory board also hoata the architect for Google Scholar, Anurag Acharya.

@@

This field is moving quickly. In December 2020, at least one journal (eLife ‘publish then review’) announced that they won’t accept a submission until a preprint has been registered. Thus all reviews are made on preprints. Other journals, like PLoS, are announcing in house preprint servers. You should expect this area to rapidly change in the coming years, so no matter when you are reading this, you are more likely than ever to need to submit a preprint.

At the time of writing, there are still some journals that make it a condition of submission that there is no preprint. Make sure you check within your target journal list.

If you choose to publish in an [overlay journal](#overlay), then you'll have to deposit your submission onto a preprint server.

@@

Peer review is held as a gold standard in scientific publishing (although a [silver standard](#fixing) would be a more realistic), and there’s certainly a lot to that. It ensures that published material has been read and its contents assessed independently. But peer review is fallible, because scientists are all humans. These problems are discussed at length in [Part IV](#problempeerreview).

@@

All of this is very good, but will people actually read and comment?
A quick look at the sites will tell you a lot about the level of reviewing that is currently going on in biosciences preprints. Looking at the top 10 articles on BioRxiv (zoology section) confirmed my suspicions. Plenty of tweets about the articles, but none of them had any comments, let alone reviews. Indeed, a further trawl through PeerJ Preprints, also found no comments.

@@

It is clear that preprints are with us to stay. The year of the COVID-19 pandemic (2020) saw an explosion of preprint papers on the topic, but also saw the misunderstanding of what these articles mean by the press and general public alike. Rapid sharing of results via preprint servers has already been put in place following the outbreak of Zika virus in South America (back in 2015-16), but the global nature of the Covid public health crisis saw much larger numbers of preprints being placed online.

@@

Another question is what happens to manuscripts that are placed on preprint servers, are then sent out for review but not published because they are fundamentally flawed? It’s not as if the reviews are not made, but there is no automatic link to the reviews by the journal that conducted them.

@@

@@

I edit for the journal [PeerJ](www.peerj.com). Although there can be various reasons to be rejected from PeerJ, normally it means that your paper is not scientifically sound. As PeerJ has no selection for impact, rejection does not normally mean that it can be simply submitted to another journal. I have noticed that manuscripts that I have rejected from PeerJ are still available as preprints without any comment on their failure during peer review. In my opinion, this is not good as it essentially ignores the input given by both reviewers and editors. The article appears as if it has had no comments or attention, when this is not the case. In a system where we move to relying more on preprints, why would we want to ignore chosen peer reviewers for whom this article was within their specialist area? According to Google Scholar, the rejected article is gaining citations, again raising concerns that rejection by peer review is not a hurdle to entering the scientific literature. All of this calls for reviews to be linked more directly to preprints, no matter where they are published. A model that deals directly with this is [overlay journals](#overlay).

@@

Having said that preprints won't replace the role of peer review, what if we did have good, editorially coordinated peer review of preprints? What if, instead of these manuscripts effectively leaving the preprint system, they were updated together with the reviews that prompted the updates, each with their own linked [DOI](#DOI)? What if the journals themselves were simply pointing to collections of papers that had been curated in this way? Simply a website that throws a veneer of a journal as waypoints to peer reviewed journals. This world has already been imagined and is functioning in mathematics, where Overlay Journals have begun to prosper.

According to Brown [-@brown2010introduction], the idea of overlaying has been with us for some time, and exists as websites that offer a series of links to other papers. In this way, a review article could be considered an 'overlay paper', the contents of Web of Science as an 'overlay database'. But, for me at least, this is not where the real potential lies. Instead, imagine the overlay journal as a way in which academics entirely remove the need for publishers. The need for this is increasingly evident as we become more familiar with the ways in which we rely on traditional publishing models to pervade our scientific project with confirmation bias. Overlay journals no longer require a publisher to store the publication. This is done at the preprint server. The reviews are housed at the same arXiv site (or would be in an ideal and transparent version)[@rittman2020fast], as is the manuscript in its final form after being accepted by the overlay journal editor. The authors themselves are responsible for the final layout. The Overlay Journal co-ordinates the reviews and conducts the editorial work, and then simply acts as a pointer to the finished product: no papers, no publishers, no editorial management software, no costs and all papers are [Diamond OA](#diamond)!

The journal [_Discrete Analysis_](https://discreteanalysisjournal.com/) (indexed in both Web of Science and Scopus) was the first of these new 'arXiv overlay journals' (since 2017), and following this link will allow you to quickly appreciate what an Overlay Journal is. Each 'published' paper still sits on its original preprint server. The overlay journal itself offers a brief editorial summary of what you'll find if you click through to the paper. This is a fantastic idea in that it pitches editors back into being responsible content curators. As an editor I'd want to be motivated to publish a paper that I liked in order to write an editorial summary about it.

Because the only the accepted version is provided with an 'article number' and the style file of the journal layout, the author then produces the final version of record (VoR) of the accepted manuscript by running the style file with LaTeX. All of this is possible with free software, for example by using R Markdown [@xie2018r].

Using preprint servers also allows the entire process to be [transparent](#transparency2), very quickly becoming associated with other great initiatives like the Centre for Open Science - [OSF](https://www.cos.io/).

@@

> They're probably only going to succeed in disciplines where a no-frills approach to publishing is acceptable... 
> I think that the real threat to our traditional model... is if Overlay Journals have **Impact Factors** and can provide the same services, and they are free... then I think that that does pose a threat.
>
> ---[Anonymous from Highwire Press Inc. 2017](https://youtu.be/ADecAXLrKoA)

@@

One of the original electronic journals, [Journal of Medical Internet Research (JMIR)](https://www.jmir.org/), announced in 2019 that it will launch an overlay journal covering biology, a so called 'superjournal', JMIRx [@eysenbach2019celebrating]. This overlay journal operates by editors choosing preprints that they want to publish ('editorial prospecting'), and then approaching authors and reviewers, and also by authors pitching their preprints to the editors. Today [JMIRx|Bio](https://bio.jmirx.org/) accepts any preprint published on [bioRxiv](#bioRxiv). Although JMRIx|Bio, and sister journal [JMRIx|Psy](https://psy.jmirx.org/), were launched in 2020, I cannot find any articles submitted (by mid-2021). The sister journal [JMIRx|Med](https://med.jmirx.org/), launched at the same time and in the same area as other JMIR journals, and has a rapidly expanding publication base.

There's an excellent tie-in here with [transparency](#transparency). Because preprints are [Diamond OA](#diamond), and reviews are OA, the process is all transparent.

A nearly 'overlay' model is [Peer Community in Evolutionary Biology](https://evolbiol.peercommunityin.org/), launched in January 2017. This comes very close to the 'arXiv Overlay Journal' model described above. These preprints are submitted to PCI-Evol Biol, and are reviewed and (if they aren't rejected), a recommendation is given. The site then publishes the recommendation from peers as well as pointing to the preprint. However, unlike _Discrete Analysis_, the preprint remains 'unpublished' despite the peer review and can then be taken onto a traditional journal. There is a  growing [list of journals](https://evolbiol.peercommunityin.org/about/about) whose editors will accept recommendations from PCI-EvolBiol, and may use the reviews when appropriate. However, it's also worth noting that there are a small number of journals that will not accept preprints recommended by PCI-EvolBiol. While Peer Community in Evolutionary Biology does not publish their peer reviewed articles, another initiative from Peer Community In takes a step backwards to get a step closer ([see below](#commitment)).

The journal [eLife](https://elifesciences.org/) is also taking steps toward becoming an Overlay Journal, with their implementation of a **preprint only** submission route [@eisen2020implementing]. Although eLife appears to embrace all the advantages of transparency in their use of preprints, there is still a significant barrier that has recently jumped from $2500 to $3000 for the privilege to publish (eLife has an APC waiver system which is not seen by editors). Again the question of what exactly scholars are paying such high fees for comes to the forefront.

### Preregistration and a commitment to publish {#commitment}
Registering your proposal (or any research plan), means that you can present a historical document to a journal (probably four to five years later) to show that you have tested the hypotheses that you originally intended to. This is simply a way of being transparent in your science, and enables you to demonstrate that you have not been [p-hacking or HARKing](#transparency). Similarly, you can demonstrate that you are not '[salami-slicing](#salami)' your results.

Another initiative from Peer Community In is the possibility of submitting to their [Registered Reports](https://rr.peercommunityin.org/), which goes much further towards removing the confirmation bias. The Registered Report (RR) is in effect the registration of a proposal (i.e. preregistration) with review. If the RR is approved by reviewers then the study is, in principle, given the green light for publication whether or not the hypothesis posed is accepted or rejected. I say “in principle” because those same reviewers are shown the manuscript again once the results are in. They need to check that the methods proposed were followed, the analyses were conducted in the same way they were proposed, and that the conclusions are justified by the results. Peer Community In are offering to organise the two sets of peer review. In addition, there are a bunch of journals that have already signed up to accept RRs that are signed off after completion (notable among these is PeerJ). To me, this represents an important step in the right direction toward [transparency](#transparency2), and the elimination of confirmation bias. What would be great to see is the number of conventional journals sign up with RRs based on the quality of the study design and execution, and the concomitant abandonment of [Impact Factor](#impactfactor) as a driving force in publishing.

### Beware of publishers' preprint servers
Publishers have a lot to be worried about preprint servers as they have the potential to take away their business. For this reason, you will see that some publishers have launched their own preprint services. Don't be lulled into submitting to the preprint server of a publisher. There are plenty of preprint servers that have nothing to do with publishers, including the most ubiquitous. These are non-profit transparent organisations, and we all have an interest in them staying that way. Simply put, there's no need to use a publishers' preprint server (this is the same principle as your data, [see here](#suppmat)).

@@

The launch in 2018 of [Peeriodicals](https://peeriodicals.com/) puts another twist on the idea of an Overlay Journal. This time, the Peeriodical is a site where anyone puts together a collection of any published papers or pre-prints that they curate themselves on a topic. They don't pass them out for any further review, but they are in effect another kind of curated Overlay.

@@

- Your manuscript needs to be cited by another that you are submitting and you are worried that the peer review process will take too long.
  + This is especially appropriate in the need to be [transparent](#transparency2) to demonstrate that you are not '[salami slicing](#salami)' your work.
- You are applying for a scholarship, a grant or job and want to be able to show that you have a body of work that is ready to be published, even if it is not formally published yet. Using preprints, you can allow the hiring committee or potential employer access to your work. This is really much more impressive than claiming manuscripts are 'in preparation' on your CV.
- You are presenting some unpublished work at a [conference](#conference) and you want people to be able to access it (e.g. through scanning a QR code)
- You are submitting a grant application and want to demonstrate that you have sufficient data, although it isn't yet published
- You are aware that another lab is working on a similar project and are worried that submitting to peer review will scoop your findings. 
- Your work has immediacy that it might not have after (potentially) 3 months of peer review. It may be that by releasing your preprint you can contribute to an ongoing debate that otherwise you’ll potentially miss.
- It's free. No APC or other fees are involved with depositing your preprint
- You have the potential to increase your network when people you have never met read your work.
- You are concerned that you’ve missed something important or perhaps analysed something in a novel way that others might be able to help with. You want this chance at feedback before submitting to peer review.
- Your manuscript crashed out of peer review with comments that you felt were unfair or unsubstantiated. You are looking for more balanced comments.
- In the above case, you might be able to use your preprint as leverage to persuade an editor that your contribution should be fast tracked into their journal.
- If you can generate enough buzz and positive feedback, you might be able to get leverage on an editor for submission to a journal with a higher impact factor.
- You have a working group that you actively want to share your publication with, even before it is published
- This is the only submission route for 'Overlay Journals'
  + [PCI-Evol Biol](##PCI-EvolBiol) in Biological Sciences
  + [JMIRx|Bio](https://bio.jmirx.org/) accepts any preprint published on [bioRxiv](#bioRxiv) since 2020
- This is the only submission route for some other journals [e.g. eLife @eisen2020implementing]

@@

- Any of your co-authors don’t want the manuscript submitted as a preprint before peer review
- You feel that public access might mean that your results are misinterpreted
 + this should be on you to get it right before you submit it
- There is a real chance that others can use the access to your work and publish it before you
 + It's worth adding here that while you might believe that there are lots of people out there who might want to steal your work, this is a general paranoia that is very common in early career researchers. Few fields in biological sciences really have valid examples of data theft or idea theft. 
 - Preprints are another example of how everything is too rushed these days.
  + I've heard this opinion, but wonder why these authors wouldn't simply hold onto their manuscript until they feel that it's ready.

>>>>

So far in this book, I have explained about the process of [peer review](#peerreview) how important this is and later I explain why it's the [silver](#fixing) (and not gold) standard in science. Preprints came to us from the world of physics. An academic world that is moving so quickly that many inside it don't want to wait for peer review before making their work public. They date back to 1991, and were the brainchild of Paul Ginsparg with his preprint server, *arXiv*. Today, *arXiv* hosts nearly 2 million articles in 8 subject areas: Physics, Mathematics, Computer Science, Quantitative Biology, Quantitative Finance, Statistics, Electrical Engineering and Systems Science and Economics.

@@

In an analysis by Abdill and Blekhman [-@abdill2019tracking] back in 2018, there were 37648 preprints uploaded to *bioRxiv*. But in the world of preprints, this information is quite dated as back then there were only \~2000 uploads each month. Luckily, these same authors regularly scrape data regarding submissions to *bioRxiv* and I'm able to reproduce these live data below. It is interesting to look back at the concern back in 2015 that preprint servers like *bioRxiv* might not catch on in the biological sciences [@vale2015accelerating], and it is still true to say that the number of preprints in the life sciences is still dwarfed by the annual number of publications, whereas physics has seen the opposite trend.

@@

What you see is a doubling of bioRxiv submissions dating back to May 2020 (Figure \@ref(fig:bioRxiv-line-annotated)), a few months after the start of the global COVID-19 pandemic. (Go to the [RXivist website](https://rxivist.org/stats) to see the incredible spike in [medRxiv](https://www.medrxiv.org/) data at the same point in time.) Many scientists had spent a few months working from home. Some had been productive, and many decided to move this productivity for the first time onto bioRxiv.

@@

The preprint revolution has not gone unnoticed by the tech giants. Back in 2017 the [Chan Zuckerberg Initiative](https://chanzuckerberg.com/) made an undisclosed donation to *bioRxiv*. The advisory board also hos the architect for Google Scholar, Anurag Acharya.

@@

This field is moving quickly. In December 2020, at least one journal (*eLife* ‘publish then review’) announced that they won’t accept a submission until a preprint has been registered. Thus all reviews are made on preprints. Other journals, like *PLoS*, are announcing in house preprint servers. You should expect this area to rapidly change in the coming years, so no matter when you are reading this, you are more likely than ever to need to submit a preprint.

At the time of writing, there are still some journals that make it a condition of submission that there is no preprint. Make sure you check within your [target journal list](#shortlist).

If you choose to publish in an [overlay journal](#overlay), then you'll have to deposit your submission onto a preprint server. But note that some preprint servers are incompatible with overlay journals as they do not allow [Version of Record](#VoR) to be placed there (see [below](#peer-community-in)).

@@

Peer review is often regarded as a gold standard in scientific publishing (although a [silver standard](#fixing) would be more realistic), and there’s certainly a lot to that. It ensures that published material has been read and its contents assessed independently. But peer review is fallible, because scientists are all humans. These problems are discussed at length in [Part IV](#problempeerreview).

@@

All of this is very good, but will people actually read and comment?
A quick look at the sites will tell you a lot about the level of reviewing that is currently going on in biosciences preprints. Looking at the top 10 articles on _BioRxiv_ (zoology section) confirmed my suspicions. Plenty of tweets about the articles, but none of them had any comments, let alone reviews. Indeed, a further trawl through PeerJ Preprints, also found no comments.

@@

It is clear that preprints are with us to stay. The year of the COVID-19 pandemic (2020) saw an explosion of preprint papers on the topic, but also saw the misunderstanding of what these articles mean by the press and general public alike. Rapid sharing of results via preprint servers has already been put in place following the outbreak of Zika virus in South America (back in 2015-16), but the global nature of the COVID-19 public health crisis saw much larger numbers of preprints being placed online.

@@

Another question is what happens to manuscripts that are placed on preprint servers, and are then sent out for review but not published because they are fundamentally flawed? It’s not as if the reviews are not made, but there is no automatic link to the reviews by the journal that conducted them.

@@

In a way, this should be a call for more journals to use preprint servers. The more that journals become involved in the curation of preprints, the more reviews that they invite will likely rest alongside the preprints themselves. If reviews that spot and reject manuscripts based on fundamental flaws rest with them, this would save a lot of time should they be sent out for review again.

@@

I edit for the journal [PeerJ](www.peerj.com). Although there can be various reasons to be rejected from PeerJ, normally it means that your paper is not scientifically sound. As PeerJ has no selection for impact, rejection does not normally mean that it can be simply submitted to another journal. I have noticed that manuscripts that I have rejected from PeerJ are still available as preprints without any comment on their failure during peer review. In my opinion, this is not good as it essentially ignores the input given by both reviewers and editors. The article appears as if they have had no comments or attention, when this is not the case. In a system where we move to relying more on preprints, why would we want to ignore chosen peer reviewers for whom this article was within their specialist area? According to Google Scholar, the rejected article is gaining citations, again raising concerns that rejection by peer review is not a hurdle to entering the scientific literature. All of this calls for reviews to be linked more directly to preprints, no matter where they are published. A model that deals directly with this is [overlay journals](#overlay).

@@

Having said that preprints won't replace the role of peer review, what if we did have good, editorially coordinated peer review of preprints? What if, instead of these manuscripts effectively leaving the preprint system, they were updated together with the reviews that prompted the updates, each with their own linked [DOI](#DOI)? What if the journals themselves were simply pointing to collections of papers that had been curated in this way? Simply a website that throws a veneer of a journal as waypoints to peer reviewed articles. This world has already been imagined and is functioning in mathematics, where Overlay Journals have begun to prosper.

According to Brown [-@brown2010introduction], the idea of overlaying has been with us for some time, and exists as websites that offer a series of links to other papers. In this way, a review article could be considered an 'overlay paper', the contents of Web of Science as an 'overlay database'. But, for me at least, this is not where the real potential lies. Instead, imagine the overlay journal as a way in which academics entirely remove the need for publishers. The need for this is increasingly evident as we become more familiar with the ways in which we rely on traditional publishing models to pervade our scientific project with confirmation bias. Overlay journals no longer require a publisher to store the publication. This is done at the preprint server. The reviews are housed at the same *arXiv* site (or would be in an ideal and transparent version) [@rittman2020fast], as is the manuscript in its final form after being accepted by the overlay journal editor. The authors themselves are responsible for the final layout. The Overlay Journal co-ordinates the reviews and conducts the editorial work, and then simply acts as a pointer to the finished product: no papers, no publishers, no editorial management software, no costs and all papers are [Diamond OA](#diamond)!

The journal [_Discrete Analysis_](https://discreteanalysisjournal.com/) (indexed in both Web of Science and Scopus) was the first of these new '*arXiv* overlay journals' (since 2017), and following this link will allow you to quickly appreciate what an Overlay Journal is. Each 'published' paper still sits on its original preprint server. The overlay journal itself offers a brief editorial summary of what you'll find if you click through to the paper. This is a fantastic idea in that it pitches editors back into being responsible content curators. As an editor I'd want to be motivated to publish a paper that I liked in order to write an editorial summary about it.

Because only the accepted version is provided with an 'article number' and the style file of the journal layout, the author then produces the final version of record (VoR) of the accepted manuscript by running the style file with LaTeX. All of this is possible with free software, for example by using R Markdown [@xie2018r].

Using preprint servers also allows the entire process to be [transparent](#transparency2), very quickly becoming associated with other great initiatives like the Centre for Open Science - [OSF](https://www.cos.io/). However, some preprint servers will not host formatted manuscripts or the [Verson of Record](#VoR) which means that if more journals switched to the overlay journal format, authors would have to be careful of exactly what restrictions their chosen preprint server has.

@@

> "They're probably only going to succeed in disciplines where a no-frills approach to publishing is acceptable... 
>
> I think that the real threat to our traditional model... is if Overlay Journals have **Impact Factors** and can provide the same services, and they are free... then I think that that does pose a threat."
>
> ---[Anonymous from Highwire Press Inc. 2017](https://youtu.be/ADecAXLrKoA)

@@

One of the original electronic journals, [*Journal of Medical Internet Research* (*JMIR*)](https://www.jmir.org/), announced in 2019 that it will launch an overlay journal covering biology, a so called 'superjournal', *JMIRx* [@eysenbach2019celebrating]. This overlay journal operates by editors choosing preprints that they want to publish ('editorial prospecting'), and then approaching authors and reviewers, and also by authors pitching their preprints to the editors. Today [*JMIRx|Bio*](https://bio.jmirx.org/) accepts any preprint published on [*bioRxiv*](#bioRxiv). Although *JMIRx|Bio*, and sister journal [*JMRIx|Psy*](https://psy.jmirx.org/), were launched in 2020, I cannot find any articles submitted (by mid-2021). The sister journal [*JMIRx|Med*](https://med.jmirx.org/), launched at the same time and in the same area as other JMIR journals, and has a rapidly expanding publication base.

Another preprint led service is [_F1000Research_](www.f1000research.com), which grew (via a buy out by Taylor & Francis) from the peer recommendation site [Faculty Opinions](www.facultyopinions.com). _F1000Research_ requires preprint submission on their server (see [cautionary note below](#beware)), and then co-ordinates postprint peer review from 5 author suggestions. For transparency, this is all accessible online and in the same location. Like other [Gold OA](#gold) journals, _F1000Research_ charges an impressive [Article Processing Charge (APC)](#gold), and does not aspire to an [Impact Factor](#noIF).

The journal [eLife](https://elifesciences.org/) is also taking steps toward becoming an Overlay Journal, with their implementation of a **preprint only** submission route [@eisen2020implementing]. Although eLife appears to embrace all the advantages of transparency in their use of preprints, there is still a significant [APC](#gold) barrier that has recently jumped from $2500 to $3000 for the privilege to publish (eLife has an [APC](#gold) waiver system which is not seen by editors). Again the question of what exactly scholars are paying such high fees for comes to the forefront.

All of these models tie-in with [transparency](#transparency). Because preprints and reviews are OA, the review process is transparent.

#### Peer Community In

*Peer Community In* deserve their own subheading in this section as they are the closest that you can get to an open publishing experience in the Biological Sciences. They drive a number of initiatives that are designed to meet those committed to open publishing, as well as those who want to test the waters.

A nearly 'overlay' model is [*Peer Community in Evolutionary Biology*](https://evolbiol.peercommunityin.org/), launched in January 2017. This comes very close to the '*arXiv* Overlay Journal' model described above. These preprints are submitted to PCI-Evol Biol, and are reviewed and (if they aren't rejected), a recommendation is given. The site then publishes the recommendation from peers as well as pointing to the preprint. However, unlike _Discrete Analysis_, the preprint remains 'unpublished' despite the peer review and can then be taken onto a traditional journal. There is a  growing [list of journals](https://evolbiol.peercommunityin.org/about/about) whose editors will accept recommendations from PCI-EvolBiol, and may use the reviews when appropriate. However, it's also worth noting that there are a small number of journals that will not accept preprints recommended by PCI-EvolBiol. While *Peer Community in Evolutionary Biology* does not publish their peer reviewed articles, another initiative from *Peer Community In* takes a step backwards to get a step closer ([see below](#commitment)).

In 2021, *Peer Community In* launched *[Peer Community Journal](https://peercommunityjournal.org/)*; a new [Diamond Open Access](#diamond) journal that publishes articles that have undergone PCI reviewing. Technically, *Peer Community Journal* claim that they are not an Overlay Journal as the journal hosts the final pdf [Version of Record](#VoR) leaving authors free to use any preprint server (see [above](#overlay)). The site relies on sponsorship to maintain the servers and archive content. Aside from this, *Peer Community Journal* achieves the highest accolade in being [Diamond Open Access](#diamond).

### Preregistration and a commitment to publish {#commitment}
Registering your proposal (or any research plan), means that you can present a historical document to a journal (probably four to five years later) to show that you have tested the hypotheses that you originally intended to. This is simply a way of being transparent in your science, and enables you to demonstrate that you have not been [p-hacking or HARKing](#transparency). Similarly, you can demonstrate that you are not '[salami-slicing](#salami)' your results. Moreover, there is some evidence to suggest that reviews of preregistered research plans inhibit researchers from leveraging their own beliefs to generate the kind of surprising results we associate with high [Impact Factors](#impactfactor) [@gross2021why].

Another initiative from _Peer Community In_ is the possibility of submitting to their [Registered Reports](https://rr.peercommunityin.org/), which goes much further towards removing the confirmation bias. The Registered Report (RR) is in effect the registration of a proposal (i.e. preregistration) with review. If the RR is approved by reviewers then the study is, in principle, given the green light for publication whether or not the hypothesis posed is accepted or rejected. I say “in principle” because those same reviewers are shown the manuscript again once the results are in. They need to check that the methods proposed were followed, the analyses were conducted in the same way they were proposed, and that the conclusions are justified by the results. Peer Community In are offering to organise the two sets of peer review. In addition, there are a bunch of journals that have already signed up to accept RRs that are signed off after completion (notable among these is PeerJ). To me, this represents an important step in the right direction toward [transparency](#transparency2), and the elimination of confirmation bias. What would be great to see is the number of conventional journals sign up with RRs based on the quality of the study design and execution, and the concomitant abandonment of [Impact Factor](#impactfactor) as a driving force in publishing.

### Beware of publishers' preprint servers {#beware}
Publishers have a lot to be worried about preprint servers as they have the potential to take away their business. For this reason, you will see that some publishers have launched their own preprint services. Don't be lured into submitting to the preprint server of a publisher. There are plenty of preprint servers that have nothing to do with publishers, including the most ubiquitous. These are non-profit transparent organisations, and we all have an interest in them staying that way. Simply put, there's no need to use a publishers' preprint server or other tools that they offer. Instead use Open Source Tools and transparent resources [@kramer2016innovations] to avoid [academic capture](#capture).

@@

The launch in 2018 of *[Peeriodicals](https://peeriodicals.com/)* puts another twist on the idea of an Overlay Journal. This time, the *Peeriodical* is a site where anyone puts together a collection of any published papers or pre-prints that they curate themselves on a topic. They don't pass them out for any further review, but they are in effect another kind of curated Overlay.

@@

- Your manuscript needs to be cited by another that you are submitting and you are worried that the peer review process will take too long.
  + This is especially appropriate in the need to be [transparent](#transparency2) to demonstrate that you are not '[salami slicing](#salami)' your work.
- You are applying for a scholarship, a grant or job and want to be able to show that you have a body of work that is ready to be published, even if it is not formally published yet. Using preprints, you can allow the hiring committee or potential employer access to your work. This is really much more impressive than claiming manuscripts are 'in preparation' on your CV.
- You are presenting some unpublished work at a [conference](#conference) and you want people to be able to access it (e.g. through scanning a QR code).
- You are submitting a grant application and want to demonstrate that you have sufficient data, although it isn't yet published.
- You are aware that another lab is working on a similar project and are worried that submitting to peer review will scoop your findings. 
- Your work has immediacy that it might not have after (potentially) 3 months of peer review. It may be that by releasing your preprint you can contribute to an ongoing debate that otherwise you’ll potentially miss.
- It's free. No APC or other fees are involved with depositing your preprint.
- You have the potential to increase your network when people you have never met read your work.
- You are concerned that you’ve missed something important or perhaps analysed something in a novel way that others might be able to help with. You want this chance at feedback before submitting to peer review.
- Your manuscript crashed out of peer review with comments that you felt were unfair or unsubstantiated. You are looking for more balanced comments.
- In the above case, you might be able to use your preprint as leverage to persuade an editor that your contribution should be fast tracked into their journal.
- If you can generate enough buzz and positive feedback, you might be able to get leverage on an editor for submission to a journal with a higher impact factor.
- You have a working group that you actively want to share your publication with, even before it is published
- This is the only submission route for 'Overlay Journals'
  + [Peer Community Journal](https://peercommunityjournal.org/)
  + [PCI-Evol Biol](##PCI-EvolBiol) in Biological Sciences.
  + [JMIRx|Bio](https://bio.jmirx.org/) accepts any preprint published on [bioRxiv](#bioRxiv) since 2020
- This is the only submission route for some other journals [e.g. _eLife_ @eisen2020implementing].

@@

- Any of your co-authors don’t want the manuscript submitted as a preprint before peer review.
- Someone wants you to deposit a preprint in a [publisher owned repository](#capture).
- Preprints can, and have, caused real damage to people when they are made in areas of public interest, such as public health. Roy and Edwards [-@roy2022addressing] provide some useful guidelines if you feel that your preprint could stray into this area.
- You feel that public access might mean that your results are misinterpreted.
  + this should be on you to get it right before you submit it.
- There is a real chance that others can use the access to your work and publish it before you.
  + It's worth adding here that while you might believe that there are lots of people out there who might want to steal your work, this is a general paranoia that is very common in early career researchers. Few fields in biological sciences really have valid examples of data theft or idea theft. 
- Preprints are another example of how everything is too rushed these days.
   + I've heard this opinion, but wonder why these authors wouldn't simply hold onto their manuscript until they feel that it's ready.

### 2021-12-15 2.02-letter.Rmd
Add files via upload
# with lots of changes following copy edits and proofs

Some journals have a text box in their submission software for you to paste the text of this letter. If there is a text box, then I suggest that you do not compose it on the fly, but plan it out and write it in a word processor so that you can catch any silly spelling or grammatical errors. Others journals might expect you to upload a pdf letter on headed paper from your institution. Make sure you know what is required and that you are prepared. The reason for a formal signed letter on headed paper is that in some submissions, this letter is used as a legal declaration.

@@

This is a moot point in that you will never know. Where there are legal requirements, it's not likely that editors will read them, but editorial managers may well make sure that these are present in order not to get a pre-review ([desk](#desk)) rejection.

Chief editors will likely read parts of the cover letter along with the abstract before deciding which Associate Editor should be assigned. Some chief editors will even leave this role to the Associate Editor. The Associate Editor should read your cover letter in full, together with the metadata and other salient information therein. They should also [read the entire manuscript](#editors).

>>>>

Some journals have a text box in their submission software for you to paste the text of this letter. If there is a text box, then I suggest that you do not compose it on the fly, but plan it out and write it in a word processor so that you can catch any silly spelling or grammatical errors. Other journals might expect you to upload a pdf letter on headed paper from your institution. Make sure you know what is required and that you are prepared. The reason for a formal signed letter on headed paper is that in some submissions, this letter is used as a legal declaration.

@@

This is a moot point in that you will never know. Some editors definitely read every cover letter [@kenar2016dear]. Even new journals have retained the need for a cover letter when they have stripped down the submission process to a minimum (e.g. _eLife_), and so my suggestion is that it is needed and, hopefully, it will be read. However, every editor at every journal is likely to have their own method of what they regard as important to read [@moustafa2015does], with many actively trying to [minimise this](#editors). Where there are legal requirements, it's not likely that editors will read them, but editorial managers may well make sure that these are present in order not to get a pre-review ([desk](#desk)) rejection.

Chief editors may read parts of the cover letter along with the abstract before deciding which Associate Editor should be assigned. Some chief editors will even leave this to the Associate Editor. The Associate Editor should read your cover letter in full, together with the metadata and other salient information therein. They should also [read the entire manuscript](#editors).  It is worth noting that reviewers are not given access to cover letters.
